What We Eval in the Shadows

### 2021-08-10 paper/main.tex
4

@@

Out of the minimized expressions, it is \c{BLOCK}, \c{<-} and
\c{F(F(X)} that contribute to the vast majority of side effects. Concretely,
92\% of \c{BLOCK}, 87\% of \c{<-} and 15\% of \c{F(F(X))} expressions contain
side effects. In the case of \c{BLOCK}, most of the side effects happen in the
environment passed to \eval (which by default is the parent environment, \ie
the local or function in our classification). For \c{<-} it is 15\%, however,
that is only because of the \c{plyr::allocate\_column} which does the side
effect in the environment containing the data frame that is the target for the
newly allocated column. Without this call, \c{<-} is similarly predictable with
96\% of side effects in the passed environment. For \c{F(F(X))} minimization
for, it is less clear as 48\% of side effects happen in a different environment
to the one that was passed to \eval.

\paragraph{Discussion} In JavaScript assignments in \eval can happen in either
local scope or less often, when called through an alias, in the global scope. In
R, given the support for first-class environments, it can happen anywhere,
making it \eval more dangerous than it already is. However, the data suggest
that first, side-effects from \eval are not as widespread as in the case of
JavaScript,\footnote{\citep{ecoop11} shows that in the \emph{Interactive}
scenario, \eval in Javascript performs, store events can reach up to 40\% of the
events, and 7\% to 8\% of the \eval do side effects in the global scope. } and
that over half of them happen in a predictable environment.

@@

\begin{figure}[h]
\begin{lstlisting}
.makeCppMethods <- function(methods, cppclassinfo, env) {
    ...
    for(what in cppMethods[! cppMethods %in% newMethods])
        methods[[what]] <- eval(substitute(
                  function(...) .CppObject$WHAT(...), list(WHAT = as.name(what))), env)
    ...
}
\end{lstlisting}
  \caption{Rcpp::.makeCppMethods}
\end{figure}

@@

\mypara{Domain-Specific Languages} The combination of laziness, metaprogramming,
first-class environments and dynamic evaluation makes R the perfect toolkit for
building DSLs. An example of this is the \code{glue} package which provides
functions for string interpolation. The code snippet below shows an example of
\code{glue} function in action.

@@

The \code{glue} function extracts snippets of R code enclosed between \c{\{} and
\c{\}}, evaluates them using \eval, and splices the result to construct the
final string.

Another example is the \code{data.table} package that defines a DSL for data
frame manipulation using the subset operator \code{[}. This subset operation is
defined as \code{DT[i, j, k]}, where expression \code{j} is used to calculate on
a subset/reorder of rows from \code{i} grouped by \code{k}. This enables compact
queries on data frames such as \code{flights[carrier == "AA", .N, by = origin]}
which will compute the number of flights done by American Airlines (\code{"AA"})
for each origin airport. At its core, \code{[} uses \code{substitute} to capture
the operand expressions, transforms them, and evaluates them using \eval in an
environment which has data frame columns available as bindings.

\mypara{Boilerplate Removal} The combination of \code{match.call} and \code{eval}
is frequently used by modeling functions to avoid boilerplate code and ease
maintenance. Consider the example shown below of function \code{survfit.formula}
from the \code{survival} package.

>>>>

% TODO do assignments change types?

@@

\paragraph{Discussion} In JavaScript assignments in \eval can happen in either
local scope or less often, when called through an alias, in the global scope. In
R, given the support for first-class environments, it can happen anywhere,
making it \eval more dangerous than it already is. However, the data suggest
that first, side-effects from \eval are not as widespread as in the case of
JavaScript,\footnote{\citep{ecoop11} shows that in the \emph{Interactive}
scenario, \eval in Javascript performs, store events can reach up to 40\% of the
events, and 7\% to 8\% of the \eval do side effects in the global scope. } and
that over half of them happen in a predictable environment.
% TODO connect the minimization group and side-effects

@@

\begin{figure}[h]
\begin{lstlisting}
.makeCppMethods <- function(methods, cppclassinfo, env) {
    cppMethods <- names(cppclassinfo@methods)
    newMethods <- names(methods)
    for(what in cppMethods[! cppMethods %in% newMethods])
        methods[[what]] <- eval(substitute(
                  function(...) .CppObject$WHAT(...), list(WHAT = as.name(what))),
                                env)
    methods
}
\end{lstlisting}
  \caption{Rcpp::.makeCppMethods}
\end{figure}

@@

\paragraph{Domain-Specific Languages} The combination of laziness, metaprogramming,
first-class environments and dynamic evaluation make R the perfect toolkit for
building DSLs. An example of this is the \code{glue} package which provides
functions for string interpolation. The code snippet below shows an example of
this.

@@

text enclosed by c{\{} and c{\}} is
evaluated using \eval as an R expression and replaced by its result.

The ability to capture the current call as a language object (\code{match.call()}),
extract the unevaluated expression from function arguments (\code{substitute()})
and \eval with first class environment makes R the toolkit for building DSLs.
There are numerous widely spread DSLs used in R. For example, \code{data.table}
package defines a DSL for data frame manipulation using the subset operator
\code{[}. This allows one to write compact queries such as \code{flights[carrier ==
  "AA", .N, by = origin]}\footnote{The \code{data.table} subset operation is
  defined as \code{DT[i, j, k]}, where \code{j} expression is used to calculate on a
  subset/reorder of rows from \code{i} grouped by \code{k}.} which will compute the
number of flights done by American Airlines for each origin airport.
Essentially, it uses \code{substitute} to capture the operand's expressions,
transforms then appropriately and finally uses \eval to execute the query.

Another example is the \code{tidyselect} package that defines a DSL for selecting
columns in a data frame powering number of data manipulation operation in the
\code{data.table} main competitor, the \code{dply} package. It uses eval to build an
interpreter for evaluating column selectors such as \code{select(df,
starts\_with("length") \& !(name:mass))} which returns all columns in \code{df}
that starts with the string \code{length} and is not between \code{name} and
\code{mass} columns.

Another instance is the \emph{symbolic differentiation support} provided by base R package \emph{stats}, with two operators \code{D} and \code{deriv}. They support arithmetic operations and functions on real numbers such as \code{sin}, on expressions built with \code{expression()} or \code{parse}.
In \emph{MCMCglmm} (Monte Carl Markov Chain Generalised Linear Mixed Models), a \code{Dtensor} object is created: it contains expressions that are derived twice (with \code{DD}). To get a concrete value, it is evaluated in  dataframe \code{mu} where symbolic variables are bound to actual numbers.
\begin{lstlisting}
 expr<-expression(beta_1 + time*beta_2+u)
 mu = data.frame(beta_1=0.5, beta_2=1, time=3, u=2.3)
 D[i]<-eval(DD(expr, name[unlist(comb.pos[i,])]), mu)
\end{lstlisting}

\paragraph{Boilerplate Removal} The combination of \code{match.call} and \code{eval}
is frequently used by modeling functions to avoid boilerplate code and ease
maintenance. Consider the example shown below of function \code{survfit.formula}
from the \code{survival} package.
